MR-PFU: A suite of R packages for energy conversion chain analysis

### 2023-07-24 paper.Rmd
Completed section on PFUAggDatabase and related packages.

Energy flows through societal energy conversion chains (ECCs)
to enable economic activity and facilitate human flourishing.
To understand economic growth and human well-being, 
the field of energy analysis evaluates ECCs
from the primary stage 
(coal, oil, natural gas) 
to the final stage
(energy purchased by consumers, such as refined petroleum and electricity),
to the useful stage
(heat, mechanical drive, and light),
and sometimes to energy services
(transport, thermal comfort, and illumination).
Societal exergy analysis (SEA), an extension of energy analysis,
quantifies ECCs as exergy, 
the mechanical work potential of energy.

@@

Historically, societal exergy analysts have
analyzed the ECCs of individual countries
using a large number of linked spreadsheets,
starting with primary and final stage data from the [IEA](https://www.iea.org/data-and-statistics/data-product/world-energy-balances)'s
world energy balances
[@Ayres:2003ec; @Serrenho:2014aa; @Brockway:2014aa; @Brockway:2015aa]. 
Data were in varying and inconsistent formats.
The authors of this paper and others in the field
wanted to expand SEA to cover all countries, but
the spreadsheet approach to data analysis was
deemed not scalable.
We created a suite of `R` packages to create a database^[Strictly speaking, we create data frames of matrices, 
not a structured database.]
of PFU data (the PFU Database)
for all countries in the world from 1960--2020.
The packages enable, for the first time, scalable SEA.
This paper describes the design of those packages and
demonstrates briefly their use.

@@

The most important decision for the design of SEA packages
involves data format.
We selected the Physical Supply-Use Table (PSUT) framework,
a matrix approach to describing energy flows from 
resource extraction to
processing stages and, ultimately, 
to final demand [@Rocco:2016; @Guevara:2017; @Heun:2018].
The PSUT framework succinctly describes the flow of energy carriers
("products" in PSUT terminology) 
among energy conversion machines 
("industries" in PSUT terminology)
with a set of six matrices as described in the following table.

@@

Further development followed from selection of the PSUT data format.
First, the matrix-based PSUT framework brought the challenge that 
different countries and years 
have varying energy energy carriers (products) and
varying energy conversion machines (industries),
meaning that PSUT matrices for different countries and years 
have differing sizes and differing row and column names.
To get around this challenge, we created the
([matsbyname](https://github.com/MatthewHeun/matsbyname/)) 
package which enables matrix mathematics 
that respects matrix row and column names, 
inserting rows or columns of `0`s when needed.
Second,
we knew it would be convenient to perform _matrix_ mathematics 
as easily as _scalar_ mathematics
in `R` data frames
using the [tidyverse](https://www.tidyverse.org) syntax. 
We developed the [matsindf](https://matthewheun.github.io/matsindf/) package 
to enable this functionality.
Finally, manipulating row and column names proved to be a challenge, 
especially for matrices in PSUT data frames, so 
we developed the [RCLabels](https://matthewheun.github.io/RCLabels/) package
for that purpose.
The table below summarizes these packages, 
all of which are generally useful and available on 
[CRAN](https://cran.r-project.org).

@@

Several calculation steps are required to create the PFU database. 
Each step is assisted by an `R` package that we created.
First, the IEA's primary- and final-stage WEEB data must be converted
to the PSUT format, 
a task completed by the [IEATools](https://github.com/MatthewHeun/IEATools/) package. 
Second, human and animal muscle work must be calculated from 
labor and FAO data following the methodology of @Steenwyk:2022ww
using the [MWTools](https://github.com/EnergyEconomyDecoupling/MWTools) package. 
Third, the IEA's primary- and final-stage ECC data 
are extended to the useful stage by
(a) allocating final stage energy to end-use machines and
(b) multiplying allocated final energy by the 
final-to-useful efficiency of each machine.
This task is accomplished by the
[PFUDatabase](https://github.com/energyeconomydecoupling/PFUDatabase/) 
and 
[Recca](https://github.com/MatthewHeun/Recca) 
packages.
Finally, ECCs must be converted from energy terms to exergy terms. 
This step is assisted by the [Recca](https://github.com/MatthewHeun/Recca) package.
The steps to create the PSUT matrices for each country and each year
are accomplished by a
[targets](https://docs.ropensci.org/targets/)
computation pipeline
available in the [PFUDatabase](https://github.com/energyeconomydecoupling/PFUDatabase/)
package.
These packages are specific to creating the PFU database 
and are summarized in the table below.

>>>>

Energy flows through societal energy conversion chains (ECCs)
to enable economic activity and facilitate human flourishing.
To understand economic growth and human well-being, 
the field of energy analysis evaluates ECCs
from the primary stage 
(resources extracted from the environment, 
such as coal, oil, natural gas) 
to the final stage
(energy purchased by consumers, 
such as refined petroleum and electricity),
to the useful stage
(energy desired by the end user, 
such as heat, motion, and light),
and sometimes to energy services
(such as thermal comfort, transport, and illumination).
Societal exergy analysis (SEA), 
an extension of energy analysis,
quantifies ECCs in exergy terms^[Exergy is 
the mechanical work potential of energy.].

@@

Historically, societal exergy analysts have
analyzed the ECCs of individual countries
using a large number of linked spreadsheets,
starting with primary- and final-stage data 
from the [IEA](https://www.iea.org/data-and-statistics/data-product/world-energy-balances)'s
world energy balances
[@Ayres:2003ec; @Serrenho:2014aa; @Brockway:2014aa; @Brockway:2015aa]. 
Data were in varying and inconsistent formats.
The authors of this paper and others in the field
wanted to expand SEA to cover all countries, but
the spreadsheet approach to SEA was
deemed not scalable.
We created a suite of `R` packages to create a database^[Strictly speaking, we create data frames of matrices, 
not a structured database.]
of PFU data (the PFU Database)
for all countries in the world from 1960--2020.
The packages enable, for the first time, scalable SEA.
This paper describes the design of those packages and
demonstrates briefly their use.

@@

The most important decision for the suite of SEA packages
involves data format.
We authors are among those who
developed the Physical Supply-Use Table (PSUT) framework,
a matrix approach to describing energy flows from 
resource extraction to
processing stages and, ultimately, 
to final demand [@Rocco:2016; @Guevara:2017; @Heun:2018].
The PSUT framework succinctly describes the flow of energy carriers
("products" in PSUT terminology) 
among energy conversion machines 
("industries" in PSUT terminology)
with a set of six matrices described in the following table.

@@

Further development followed from selection of the PSUT framework.
First, the matrices of the PSUT framework carry the challenge that 
different countries and years 
have varying energy carriers (products) and
varying energy conversion machines (industries),
meaning that PSUT matrices for different countries and years 
have differing sizes and differing row and column names.
To get around this challenge, we created the
[matsbyname](https://github.com/MatthewHeun/matsbyname/) [@Heun-matsbyname:2023]
package which enables matrix mathematics 
that respects matrix row and column names, 
inserting rows or columns of `0`s when needed.
Second,
we knew it would be convenient to perform _matrix_ mathematics 
as easily as _scalar_ mathematics
in `R` data frames
using the [tidyverse](https://www.tidyverse.org) syntax [@Wickham:2019]. 
We developed the [matsindf](https://matthewheun.github.io/matsindf/) package [@Heun-matsindf:2023]
to enable this functionality.
Finally, manipulating row and column names proved to be a challenge, 
especially for matrices in PSUT data frames, so 
we developed the [RCLabels](https://matthewheun.github.io/RCLabels/) package [@Heun-RCLabels:2023]
for that purpose.
The table below summarizes these packages, 
all of which are generally useful and available on 
[CRAN](https://cran.r-project.org).

@@

Several calculation steps are required to create the PFU database. 
Each step is assisted by an `R` package that we created.
First, the IEA's primary- and final-stage WEEB data must be converted
to the PSUT format, 
a task completed by the [IEATools](https://github.com/MatthewHeun/IEATools/) package [@Heun-IEATools:2023]. 
Second, human and animal muscle work must be calculated from 
labor and FAO data following the methodology of @Steenwyk:2022ww
using the [MWTools](https://github.com/EnergyEconomyDecoupling/MWTools) package [@Marshall:2023ab]. 
Third, the IEA's primary- and final-stage ECC data 
are extended to the useful stage by
(a) allocating final stage energy to end-use machines and
(b) multiplying allocated final energy by the 
final-to-useful efficiency of each machine.
This task is accomplished by the
[Recca](https://github.com/MatthewHeun/Recca) [@Heun-Recca:2023]
and 
[PFUDatabase](https://github.com/energyeconomydecoupling/PFUDatabase/) [@Heun-PFUDatabase:2023]
packages.
Finally, ECCs must be converted from energy terms to exergy terms. 
This step is assisted by the [Recca](https://github.com/MatthewHeun/Recca) package.
The steps to create the PSUT matrices for each country and each year
are accomplished by a
[targets](https://docs.ropensci.org/targets/) [@Landau:2021aa]
computation pipeline
available in the [PFUDatabase](https://github.com/energyeconomydecoupling/PFUDatabase/)
package.

A second [targets](https://docs.ropensci.org/targets/) pipeline 
in the [PFUAggDatabase](https://github.com/EnergyEconomyDecoupling/PFUAggDatabase) package [@Heun-PFUAggDatabase:2023]
aggregates ECCs
by region 
(continents and world),
by product category 
(e.g., Coal and coal products, Oil and oil products, Low-temperature heat, etc.), 
by end use categories
(e.g., Residential, Transport, etc.), and
to primary, final, and useful stages.
Furthermore, 
[PFUAggDatabase](https://github.com/EnergyEconomyDecoupling/PFUAggDatabase)
calculates aggregated efficiencies.
The [Recca](https://github.com/MatthewHeun/Recca) package is used extensively by the
[PFUAggDatabase](https://github.com/EnergyEconomyDecoupling/PFUAggDatabase) 
pipeline.

The packages in the following table are specific to creating the PFU database 
and are available on GitHub.
